
\section{Introduction}

Hospitals produce a huge amount of clinical notes that have been solely used for archiving purposes and have generally been inaccessible to researchers. 
However, application of recent \acrshort{nlp} technology can make accessible the hidden knowledge of archived records, thus boosting medical research. 
An example is the English cTAKES system \cite{Savova2010}, which can recognize important medical concepts in clinical free-text documents. 
Beside extracting diseases, symptoms and treatments, the tool is also able to identify various relations between them.
This automatically extracted structured knowledge can be e.g. used to 
\begin{itemize}
	\item create family histories from medical records,
	\item find document discrepancies,
	\item spot similar cases and
	\item build question answering or semantic search systems.
\end{itemize}
While developing text processing tools for medicians is an emerging field in many developed countries, less-resourced languages lack such resources.

To be able to extract information from medical texts, they must be preprocessed properly. 
Firstly, adequate text segmentation methods 
%\footnote{While the term \emph{text segmentation} is widely used for diverse tasks, in our work it denotes the process of dividing text into tokens and sentences.} 
are required for finding token and sentence boundaries. 
Secondly, morphological tagging is an indispensable step for information extraction scenarios. 
Considering the case of Hungarian, there are only a few studies on processing medical records. 
Recently, Siklósi et al. \cite{Siklosi2012,Siklosi2013} have presented a system that is able to correct spelling errors in clinical notes. 
Their system uses a mixture of language models to generate correction candidates, however it focuses only on correctly segmented words. 
Beside error correction, an abbreviation resolution method was also presented by them \cite{Siklosi2013b}, however, problems of text segmentation and  morpho-syntactic tagging are still untouched. 
Furthermore, as far as we know, no study investigates such preprocessing tasks on Hungarian clinical texts. 

Therefore, this chapter presents accurate preprocessing algorithms for noisy medical texts.
Methods were developed and presented for only Hungarian, but they are designed in a way to perform well on other morphologically complex languages as well. 
Firstly, an effective method is introduced for detecting sentence and token boundaries.
The presented system builds on well-known tokenization rules boosting them with the knowledge of a morphological analyzer and the output of an unsupervised filtering algorithm.
Secondly, tagging experiments are presented yielding a viable morphological tagger for Hungarian electronic health records. 
The proposed tool builds on PurePos fixing its most common errors regarding the domain.

\section{Segmenting texts of electronic health records}\label{sec:clin_segm}
\input{Clinical/section_segmentation}

\pagebreak

\section{Morphological tagging of clinical notes}\label{sec:clin_tag}
\input{Clinical/section_tagging}

