\begin{chapquote}{Psalms 121:2}
``My help will come from the Lord, who made heaven and earth.''
\end{chapquote}

First of all, I would like to say thank you to my scientific advisor, Gábor Prószéky for guiding and supporting me over the years.

I am also thankful to Attila Novák for the fruitful conversations and his useful advices.
I would like to give thanks to Nóra Wenszky as well for polishing my English and helping me to refine this study.

Conversations, launches and the always cheerful coffee breaks with my colleagues are greatly acknowledged. 
Thanks to Laci Laki, Bori Siklósi, Balázs Indig, Kinga Mátyus collaborating in numerous valuable studies.
I am also thankful to members of room 314 Pisti Endrédy, Győző Yang Zijian, Bálint Sass, Marci Miháltz, András Simonyi and Károly Varasdi for the friendly and intellectual atmosphere.

I am also grateful to the Pázmány Péter Catholic University and the MTA-PPKE Hungarian Language Technology Research Group, where I spent my PhD years. 
Thanks are due to current and former leaders Tamás Roska, Judit Nyékyné Gaizler, Péter Szolgay giving me the opportunity to conduct my research.
I would like to give special thanks to Kati Hubay and Lívia Adorján for organizing our conference trips.

Work covered in thus dissertation was supported partly by the TÁMOP 4.2.1.B -- 11/2/KMR-2011–0002 and 4.2.2/B -- 10/1–2010–0014 projects.

Most importantly, I am thankful to my family.
I cannot express enough thanks to my loving wife Jucus for tolerating my  absence and encouraging me over the years.
I am grateful to my parents and brother Tomi for their continuous support during my studies.
